Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-12-20 paper/paper_no_event.tex
Paper: fix nu->eta in Eq. 14
Minor corrections to text. Updated references.

For the analysis of GW170817 in Refs.~\cite{BNSPE, O2Catalog}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, TEOBResumSupdate}, that uses a different point-particle and spin prescription. For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency-domain models, parameter estimation results are still possible. In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. 
%Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems. It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters. 
The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (Fig. 9 of Ref.~\cite{O2Catalog}). However, a direct comparison using identical waveforms has not yet been published for BNS systems.
As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

@@

For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order. In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes. This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}. Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order. An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN, see App.~\ref{sec:taylorf2}. They are also present in the IMRPhenomPv2+NRTides and SEOBNRv4ROM+NRTides models. TEOBResumS was recently updated~\cite{TEOBResumSupdate} to include these next-to-leading terms.}.

@@

In the waveform, at the leading 2PN order, the only contribution to consider will be in the mode $h_{22}$. The total leading-order spin-squared contribution, for aligned-spin circular orbits, is given by~\cite{Kidder:1995zr, Will:1996zj, Buonanno:2012rv}
\be
h_{22}^\tr{LO-SS}=-\frac{8\nu M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A \,,
\ee
with $\Phi$ the orbital phase, $\omega = \dot{\Phi}$ the orbital frequency and $R$ the distance to the observer. Here, for a BBH,
\be\label{eq:ABBH}
\mc A^\mr{BBH}=\frac{1}{M^2}\left(\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\frac{S_2^2}{m_2^2}\right) \,,
\ee
while for a BNS, this should be replaced with
\be\label{eq:ABNS}
\mc A^\mr{BNS}=\frac{1}{M^2}\left(\kappa_1\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\kappa_2\frac{S_2^2}{m_2^2}\right) \,.
\ee
The total spin-squared contribution to the flux also includes squares of spin-orbit terms, and reads~\cite{Kidder:1995zr, Will:1996zj, Poisson1998}
\be
\mc F^\tr{LO-SS}=\frac{32\eta^2}{5} v^{14} \left[2\mc A+\frac{1}{16M^2}\left(\frac{S_1}{m_1}-\frac{S_2}{m_2}\right)^2\,\right] \,,
\ee
with $v=(M\omega)^{1/3}$.

When translating the above for the EOB factorized waveform (see e.g. (17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors and phase contribution would enter only at higher order, so that the only modification to consider is in the $\rho_{\ell m}$ factor for $\ell = 2$, $m = 2$. We have simply
\be\label{eq:rho22}
	\rho_{22}^{\textrm{LO-SS}} = \frac{1}{2}\mathcal{A} v^{4} \,,
\ee
with $\mathcal{A}$ the quantity given by~\eqref{eq:ABBH} for a BBH or~\eqref{eq:ABNS} for a BNS.

@@

The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node. The rest of the time is spent resampling the waveform with spline interpolation. For a starting frequency of 20~Hz, this takes 0.13~s when the waveform is matched with data uniformly sampled at 4096Hz. For current parameter estimation codes, this is sufficient. However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017} or relative binning~\cite{Zackay:2018qdy}. Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.

The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model. For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, and constructing a surrogate of the residual. Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger. Current state-of-the-art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set. If the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.

>>>>

For the analysis of GW170817 in Refs.~\cite{BNSPE, O2Catalog}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT), or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, NagarMessinaRettegno2018}, that uses a different point-particle and spin prescription. For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency-domain models, parameter estimation results are still possible. In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. 
%Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems. It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters. 
The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (see Fig. 9 of Ref.~\cite{O2Catalog}). However, a direct comparison using identical waveforms has not yet been published for BNS systems.
As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

@@

For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order. In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes. This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}. Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order. An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN (see Appendix~\ref{sec:taylorf2}). They are also present in the PhenomPNRT and SEOBNRT models. TEOBResumS was recently updated to include these next-to-leading terms~\cite{NagarMessinaRettegno2018}.}.

@@

In the waveform, at the leading 2PN order, the only contribution to consider will be in the mode $h_{22}$. The total leading-order spin-squared contribution, for aligned-spin circular orbits, is given by~\cite{Kidder:1995zr, Will:1996zj, Buonanno:2012rv}
\be
h_{22}^\tr{LO-SS}=-\frac{8\eta M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A \,,
\ee
with $\Phi$ the orbital phase, $\omega = \dot{\Phi}$ the orbital frequency and $R$ the distance to the observer. Here, for a BBH,
\be\label{eq:ABBH}
\mc A^\mr{BBH}=\frac{1}{M^2}\left(\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\frac{S_2^2}{m_2^2}\right) \,,
\ee
while for a BNS, this should be replaced with
\be\label{eq:ABNS}
\mc A^\mr{BNS}=\frac{1}{M^2}\left(\kappa_1\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\kappa_2\frac{S_2^2}{m_2^2}\right) \,.
\ee
The total spin-squared contribution to the flux also includes squares of spin-orbit terms, and reads~\cite{Kidder:1995zr, Will:1996zj, Poisson1998}
\be
\mc F^\tr{LO-SS}=\frac{32\eta^2}{5} v^{14} \left[2\mc A+\frac{1}{16M^2}\left(\frac{S_1}{m_1}-\frac{S_2}{m_2}\right)^2\,\right] \,,
\ee
with $v=(M\omega)^{1/3}$.

When translating the above for the EOB factorized waveform (see e.g. (17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors, and phase contribution would enter only at higher order, so that the only modification to consider is in the $\rho_{\ell m}$ factor for $\ell = 2$, $m = 2$. We have simply
\be\label{eq:rho22}
	\rho_{22}^{\textrm{LO-SS}} = \frac{1}{2}\mathcal{A} v^{4} \,,
\ee
with $\mathcal{A}$ the quantity given by~\eqref{eq:ABBH} for a BBH or~\eqref{eq:ABNS} for a BNS.

@@

The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node. The rest of the time is spent resampling the waveform with spline interpolation. For a starting frequency of 20~Hz, this takes 0.13~s when the waveform is matched with data uniformly sampled at 4096Hz. For current parameter estimation codes, this is sufficient. However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}, or relative binning~\cite{Zackay:2018qdy}. Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.

The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model. For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, then constructing a surrogate of the residual. Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger. Current state-of-the-art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set. If the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
